Flow-Based Cost Allocation in Linear Optimized Power Systems are non-unique but localizable

### 2020-07-15 paper/flow_based.tex
introduce snakemake workflow
include storages!

\addbibresource{main.bib}
\graphicspath{{../figures/}}

@@

% general symbols
\newcommand{\generation}{g_{s,t}}
\newcommand{\generationpotential}{\bar{g}_{s,t}}
\newcommand{\generationshare}[1][n]{\omega_{#1,s,t}}
\newcommand{\nodalgeneration}[1][n]{g_{#1,t}}
\newcommand{\capacityGeneration}{G_{s}}
\newcommand{\capacityGenerationUpper}{\bar{G}_{s}}
\newcommand{\capacityGenerationLower}{\ubar{G}_{s}}
\newcommand{\capacityFlow}{F_{\ell}}
\newcommand{\capacityFlowUpper}{\bar{F}_{\ell}}
\newcommand{\capacityFlowLower}{\ubar{F}_{\ell}}
\newcommand{\capexGeneration}{c_{s}}
\newcommand{\capexFlow}{c_{\ell}}
\newcommand{\opexGeneration}{o_{s}}
\newcommand{\demand}[1][n]{d_{#1,t}}
\newcommand{\nodaldemand}[1][n]{d_{#1,t}}
\newcommand{\demandshare}[1][n]{\omega_{#1,a,t}}
\newcommand{\utility}{U_{n,a,t}}
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\incidenceGenerators}[1][n]{K_{#1,s}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\ptdfEqual}[1][n]{\ptdf[#1]^\circ}
\newcommand{\slackflow}{k_{\ell}}
\newcommand{\slack}[1][n]{k_{#1}}
\newcommand{\slackk}[1][n]{k^*_{#1}}
\newcommand{\Slack}{k_{m,n}}
\newcommand{\Slackk}{k^*_{m,n}}
\newcommand{\mulowergeneration}{\ubar{\mu}_{s,t}}
\newcommand{\muuppergeneration}{\bar{\mu}_{s,t}}
\newcommand{\mulowerflow}{\ubar{\mu}_{\ell,t}}
\newcommand{\muupperflow}{\bar{\mu}_{\ell,t}}
\newcommand{\lmp}[1][n]{\lambda_{#1,t}}
\newcommand{\flow}{f_{\ell,t}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}
\newcommand{\injection}{p_{n,t}}
\newcommand{\netconsumption}[1][n]{p^{-}_{#1,t}}
\newcommand{\netproduction}[1][n]{p^{+}_{#1,t}}
\newcommand{\selfconsumption}[1][n]{p^{\circ}_{#1,t}}

@@

% \subsection{LMP independent Cost  Terms}
% 
% Example: \\
% Constraining the capacities $\capacityGeneration$  for a subset $S$ of generators to lower or upper limits in the form of 
% \begin{align}
% \capacityGeneration \ge \capacityGenerationLower \hpad \forall{n,s \in S} \\
% \capacityGeneration \le \capacityGenerationUpper \hpad \forall{n,s \in S}
% \end{align}
% or doing likewise with a subset $L$ of transmission lines,
% \begin{align}
% \capacityFlow \ge \capacityFlowLower \hpad \forall{\ell \in \textit{L}} \\
% \capacityFlow \le \capacityFlowUpper \hpad \forall{\ell \in \textit{L}}
% \end{align}
% does not have any effect on the above presented allocation in \cref{eq:lmp_allocation}. The additional cost directly translate to the Quality of Supply $\muuppergeneration$ for production and $\muupperflow - \mulowerflow$ for transmission respectively. This also counts for limiting the overall production capacity $\sum_{n,s}\capacityGeneration$ or transmission capacity $\sum_\ell \capacityFlow$. However it will distort the no profit rule for generators \cref{eq:no_profit_capex_generation} and transmission \cref{eq:no_profit_capex_flow} respectively. As shown in \cite{brown_decreasing_2020}

@@

\section{Appendix}
\subsection{LMP from Optimization}
The nodal balance constraint ensures that the amount of power that flows into a bus equals the power that flows out of a bus, thus reflects the Kirchhoff Current Law (KCL)
\begin{align}
    \sum_l \incidence \, \flow  - \nodalgeneration + \nodaldemand &= 0 \resultsin{\lmp} \Forall{n,t}
    \label[constraint]{eq:nodal_balance_lin}
\end{align}
Its shadow price mirrors the Locational Marginal Prizes (LMP) $\lmp$ per bus and time step. In a power market this is the \euro/\megawatthour-price which a consumer has to pay. Note that the flow $\flow$ in \cref{eq:nodal_balance_lin} is a passive variable only, given by \cref{eq:flow_from_ptdf}.\\

@@

As stated in \cite{brown_decreasing_2020} the capical cost of a generator $s$ is equal to the shadowprice $\muuppergeneration$ times the upper availibility $\generationpotential$ of all time steps 
\begin{align}
 \capexGeneration  = \sum_t \muuppergeneration \, \generationpotential 
 \label{eq:capital_price_generation_sum}
\end{align}
\Cref{eq:upper_generation_capacity_constraint,eq:lower_generation_capacity_constraint} which yield the KKT variables $\muuppergeneration$ and $\mulowergeneration$ imply the complementary slackness,
\begin{align}
\muuppergeneration \left( \generation - \generationpotential \, \capacityGeneration \right)  &= 0  \Forall{n,s,t} 
\label{eq:complementary_slackness_upper_generation} \\
\mulowergeneration  \, \generation &= 0 \Forall{n,s,t}
\label{eq:complementary_slackness_lower_generation} 
\end{align}
Multiplying both sides of \cref{eq:complementary_slackness_upper_generation} with $\capacityGeneration$ and using \cref{eq:capital_price_generation_sum} leads to 
\begin{align}
 \capexGeneration \, \capacityGeneration  = \sum_t \muuppergeneration \, \generation 
 \label{eq:capital_price_generation_sum}
\end{align}

@@

The yielding KKT variables $\muupperflow$ and $\mulowerflow$ are only non-zero if $\flow$ is limited by the trasmission capacity in positive or negative direction, i.e. \cref{eq:upper_flow_capacity_constraint} or \cref{eq:lower_flow_capacity_constraint} are binding. The complementary slackness 
\begin{align}
\muupperflow \left( \flow - \capacityFlow \right)  &= 0 \Forall{\ell,t}
\label{eq:complementary_slackness_upper_flow} \\
\mulowerflow \left( \flow - \capacityFlow \right) &=  0 \Forall{\ell,t}
\label{eq:complementary_slackness_lower_flow} 
\end{align}
set the respective KKT for flows staying below the thermal limit to zero.

\subsection{Full Lagrangian}
\label{sec:full_lagrangian}
\begin{align}
\notag
& \lagrangian\left(\generation, \flow, \capacityGeneration, \capacityFlow, \boldsymbol{\lambda}, \boldsymbol{\mu} \right)   =   \\  
\notag
& \;\;\; \sum_{n,s} \capexGeneration \capacityGeneration + \sum_{n, s, t} \opexGeneration \generation + \sum_{\ell} \capexFlow \, \capacityFlow  \\
\notag
&+ \sum_{n,t} \lmp \left(\sum_\ell \incidence \, \flow  - \sum_s \generation + \sum_a \demand  \right)  \\ 
\notag
&+ \sum_{\ell,c,t} \cycleprice \, \cycle \, \impedance \, \flow  \\
\notag
&+ \sum_{n,s,t} \muuppergeneration \left( \generation - \generationpotential \capacityGeneration \right)  - \sum_{n,s,t} \mulowergeneration \generation  \\
&+ \sum_{\ell,t} \muupperflow \left( \flow - \capacityFlow \right) - \sum_{\ell,t} \mulowerflow \left( \flow + \capacityFlow \right)     
\label{eq:full_lagrangian}
\end{align}
% 
where $\boldsymbol{\lambda} = \left\lbrace \lmp, \cycleprice \right\rbrace $ and $\boldsymbol{\mu} = \left\lbrace \muuppergeneration, \mulowergeneration, \muupperflow, \mulowerflow \right\rbrace $ denote the set of related KKT variables. The global maximum of the Lagrangian requires stationarity with respect to all variables. The stationarity of the generation capacity variable leads to 
\begin{align}
\pdv{\lagrangian}{\capacityGeneration}  = 0 \,\, \rightarrow \,\, 
\capexGeneration =  \sum_t \muuppergeneration \, \generationpotential  \Forall{n,s}
\label{eq:capexGeneration_duality}
\end{align}

and the stationarity of the generation to 
\begin{align}
\pdv{\lagrangian}{\generation} &= 0 \,\, \rightarrow \,\,  
\opexGeneration =  \lmp - \muuppergeneration + \mulowergeneration \Forall{n,s} \label{eq:opex_duality}
\end{align}

@@

\subsection{Example: Power Flow Allocations of different Types}
\label{sec:example_plots}

\begin{figure}[h!]
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus1_net_ebe.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus1_net_ebe}
    \end{subfigure}
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus2_net_ebe.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus2_net_ebe}
    \end{subfigure}
    \caption{Power allocations for bus~1 (a) and bus~2 (b) of the example network in \cref{fig:example_network} using equivalently allcated net power injections (scheme \ref{net}\ref{ebe}). Bus~1 retrieves 60~MW from itself and nothing from bus~2. The latter in turn retrieves 40~MW from bus~1 and self supplies 50~MW. The P2P trades are less in number and more intuitive then with allocating gross flow.}
    \label{fig:example_allocation_net_ebe}
\end{figure}

>>>>

\addbibresource{main.bib}
\graphicspath{{../figures/}{../figures/example/}}

@@

% general symbols
\newcommand{\generation}{g_{s,t}}
\newcommand{\generationpotential}{\bar{g}_{s,t}}
\newcommand{\generationshare}[1][n]{\omega_{#1,s,t}}
\newcommand{\nodalgeneration}[1][n]{g_{#1,t}}
\newcommand{\capacityGeneration}{G_{s}}
\newcommand{\capacityGenerationUpper}{\bar{G}_{s}}
\newcommand{\capacityGenerationLower}{\ubar{G}_{s}}
\newcommand{\capacityFlow}{F_{\ell}}
\newcommand{\capacityFlowUpper}{\bar{F}_{\ell}}
\newcommand{\capacityFlowLower}{\ubar{F}_{\ell}}
\newcommand{\capexGeneration}{c_{s}}
\newcommand{\capexFlow}{c_{\ell}}
\newcommand{\opexGeneration}{o_{s}}
\newcommand{\demand}[1][n]{d_{#1,t}}
\newcommand{\nodaldemand}[1][n]{d_{#1,t}}
\newcommand{\demandshare}[1][n]{\omega_{#1,a,t}}
\newcommand{\utility}{U_{n,a,t}}
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\incidenceGenerators}[1][n]{K_{#1,s}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\ptdfEqual}[1][n]{\ptdf[#1]^\circ}
\newcommand{\slackflow}{k_{\ell}}
\newcommand{\slack}[1][n]{k_{#1}}
\newcommand{\slackk}[1][n]{k^*_{#1}}
\newcommand{\Slack}{k_{m,n}}
\newcommand{\Slackk}{k^*_{m,n}}
\newcommand{\mulowergeneration}{\ubar{\mu}_{s,t}}
\newcommand{\muuppergeneration}{\bar{\mu}_{s,t}}
\newcommand{\muuppergenerationnom}{\bar{\mu}^\text{nom}_{s}}
\newcommand{\mulowergenerationnom}{\ubar{\mu}^\text{nom}_{s}}
\newcommand{\mulowerflow}{\ubar{\mu}_{\ell,t}}
\newcommand{\muupperflow}{\bar{\mu}_{\ell,t}}
\newcommand{\muupperflownom}{\bar{\mu}^\text{nom}_{\ell}}
\newcommand{\mulowerflownom}{\ubar{\mu}^\text{nom}_{\ell}}
\newcommand{\lmp}[1][n]{\lambda_{#1,t}}
\newcommand{\flow}{f_{\ell,t}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}
\newcommand{\injection}{p_{n,t}}
\newcommand{\netconsumption}[1][n]{p^{-}_{#1,t}}
\newcommand{\netproduction}[1][n]{p^{+}_{#1,t}}
\newcommand{\selfconsumption}[1][n]{p^{\circ}_{#1,t}}

@@

@@

\section{Appendix}
\subsection{LMP from Optimization}
The nodal balance constraint ensures that the amount of power that flows into a bus equals the power that flows out of a bus, thus reflects the Kirchhoff Current Law (KCL)
\begin{align}
    \sum_l \incidence \, \flow  - \nodalgeneration + \nodaldemand &= 0 \resultsin{\lmp} \Forall{n,t}
    \label[constraint]{eq:nodal_balance_lin}
\end{align}
Its shadow price mirrors the Locational Marginal Prizes (LMP) $\lmp$ per bus and time step. In a power market this is the \euro/\megawatthour-price which a consumer has to pay.\\

\subsection{Full Lagrangian}
\label{sec:full_lagrangian}
\begin{align}
\notag
& \lagrangian\left(\generation, \flow, \capacityGeneration, \capacityFlow, \boldsymbol{\lambda}, \boldsymbol{\mu} \right)   =   \\  
\notag
& \;\;\; \sum_{n,s} \capexGeneration \capacityGeneration + \sum_{n, s, t} \opexGeneration \generation + \sum_{\ell} \capexFlow \, \capacityFlow  \\
\notag
&+ \sum_{n,t} \lmp \left(\sum_\ell \incidence \, \flow  - \sum_s \incidenceGenerators\, \generation +  \demand  \right)  \\ 
\notag
&+ \sum_{\ell,c,t} \cycleprice \, \cycle \, \impedance \, \flow  \\
\notag
&+ \sum_{n,s,t} \muuppergeneration \left( \generation - \generationpotential \capacityGeneration \right)  - \sum_{n,s,t} \mulowergeneration \generation  \\
&+ \sum_{\ell,t} \muupperflow \left( \flow - \capacityFlow \right) - \sum_{\ell,t} \mulowerflow \left( \flow + \capacityFlow \right)     
\label{eq:full_lagrangian}
\end{align}
% 
where $\boldsymbol{\lambda} = \left\lbrace \lmp, \cycleprice \right\rbrace $ and $\boldsymbol{\mu} = \left\lbrace \muuppergeneration, \mulowergeneration, \muupperflow, \mulowerflow \right\rbrace $ denote the set of related KKT variables. The global maximum of the Lagrangian requires stationarity with respect to all variables. The stationarity of the generation capacity variable leads to 
\begin{align}
\pdv{\lagrangian}{\capacityGeneration}  = 0 \,\, \rightarrow \,\, 
\capexGeneration =  \sum_t \muuppergeneration \, \generationpotential  \Forall{n,s}
\label{eq:capex_generation_duality}
\end{align}

and the stationarity of the generation to 
\begin{align}
\pdv{\lagrangian}{\generation} &= 0 \,\, \rightarrow \,\,  
\opexGeneration =  \incidenceGenerators \, \lmp - \muuppergeneration + \mulowergeneration \Forall{n,s} \label{eq:opex_duality}
\end{align}

\begin{align}
0 &= \pdv{\lagrangian}{\flow}  \\ 
0 &= \sum_n \incidence \lmp  + \cycleprice \cycle \impedance  - \muupperflow + \mulowerflow \Forall{n,s} \label{eq:opex_flow_duality}
\end{align}

@@

As stated in \cite{brown_decreasing_2020} the capical cost of a generator $s$ is equal to the shadowprice $\muuppergeneration$ times the upper availibility $\generationpotential$ of all time steps 
\begin{align}
 \capexGeneration  = \sum_t \muuppergeneration \, \generationpotential 
 \label{eq:capital_price_generation_sum}
\end{align}
\Cref{eq:upper_generation_capacity_constraint,eq:lower_generation_capacity_constraint} which yield the KKT variables $\muuppergeneration$ and $\mulowergeneration$ imply the complementary slackness,
\begin{align}
\muuppergeneration \left( \generation - \generationpotential \, \capacityGeneration \right)  &= 0  \Forall{n,s,t} 
\label{eq:complementary_slackness_upper_generation} \\
\mulowergeneration  \, \generation &= 0 \Forall{n,s,t}
\label{eq:complementary_slackness_lower_generation} 
\end{align}
Multiplying both sides of \cref{eq:capex_generation_duality} with $\capacityGeneration$ and using \cref{eq:complementary_slackness_upper_generation} leads to 
\begin{align}
 \capexGeneration \, \capacityGeneration  = \sum_t \muuppergeneration \, \generation 
 \label{eq:capital_price_generation_sum}
\end{align}
The zero-profit rule for generators is obtained by multiplying \cref{eq:opex_duality} with $\generation$ and using \cref{eq:complementary_slackness_lower_generation,eq:capital_price_generation_sum} which results in 
\begin{align}
  \capexGeneration \, \capacityGeneration + \sum_t \opexGeneration \generation = \sum_t \lmp \incidenceGenerators \generation
\end{align}
It states that over the whole time span, all OPEX and CAPEX for generator $s$ (left hand side) are payed back by its revenue (right hand side).

@@

The yielding KKT variables $\muupperflow$ and $\mulowerflow$ are only non-zero if $\flow$ is limited by the trasmission capacity in positive or negative direction, i.e. \cref{eq:upper_flow_capacity_constraint} or \cref{eq:lower_flow_capacity_constraint} are binding. For flows below the thermal limit, the complementary slackness 
\begin{align}
\muupperflow \left( \flow - \capacityFlow \right)  &= 0 \Forall{\ell,t}
\label{eq:complementary_slackness_upper_flow} \\
\mulowerflow \left( \flow - \capacityFlow \right) &=  0 \Forall{\ell,t}
\label{eq:complementary_slackness_lower_flow} 
\end{align}
sets the respective KKT to zero. 
When multipying \cref{eq:capexFlow_duality} with $\capacityFlow$ and using the complementary slackness \cref{eq:complementary_slackness_upper_flow,eq:complementary_slackness_lower_flow} we obtain 
\begin{align}
 \capexFlow \, \capacityFlow = \sum_t \left( \muupperflow - \mulowerflow \right)  \, \flow
 \label{eq:capital_price_flow_sum}
\end{align}
Againg we can use this to formulate the zero-profit rule for transmission lines. We mulitply \cref{eq:opex_flow_duality} with $\flow$, which finally leads us to 
\begin{align}
\capexFlow \, \capacityFlow = \sum_n \incidence\, \lmp\, \flow + \cycleprice\, \cycle\, \impedance\, \flow 
\end{align}
It states that the congestion revenue of a line (first term right hand side) substracted by the cost for cycle constraint exactly matches its CAPEX.

@@

\subsection{Example: Power Flow Allocations of different Types}
\label{sec:example_plots}
\begin{figure}[h!]
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus1_net_ebe.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus1_net_ebe}
    \end{subfigure}
    \begin{subfigure}[c]{\linewidth}
    \includegraphics[width=\linewidth]{example_allocation_bus2_net_ebe.png}
    \vspace{-40pt}
    \subcaption{}
    \label{fig:example_allocation_bus2_net_ebe}
    \end{subfigure}
    \caption{Power allocations for bus~1 (a) and bus~2 (b) of the example network in \cref{fig:example_network} using equivalently allcated net power injections (scheme \ref{net}\ref{ebe}). Bus~1 retrieves 60~MW from itself and nothing from bus~2. The latter in turn retrieves 40~MW from bus~1 and self supplies 50~MW. The P2P trades are less in number and more intuitive then with allocating gross flow.}
    \label{fig:example_allocation_net_ebe}
\end{figure}

\subsection{Brownfield Optimization and Capacity Restricitions}

Constraining the capacities $\capacityGeneration$  for a subset $S$ of generators to lower or upper limits in the form of
\begin{align}
\capacityGeneration \ge \capacityGenerationLower \resultsin{\mulowergenerationnom} \hpad \forall{n,s \in S} \\
\capacityGeneration \le \capacityGenerationUpper \resultsin{\muuppergenerationnom} \hpad \forall{n,s \in S}
\end{align}
% or doing likewise with a subset $L$ of transmission lines,
% \begin{align}
% \capacityFlow \ge \capacityFlowLower \resultsin{\mulowerflownom} \hpad \forall{\ell \in \textit{L}} \\
% \capacityFlow \le \capacityFlowUpper \resultsin{\muupperflownom} \hpad \forall{\ell \in \textit{L}}
% \end{align}
alters the objective value as soon as one of those become bounding. 
The complementary slackness state is 
\begin{align}
 \muuppergenerationnom \left( \capacityGeneration - \capacityGenerationUpper \right) =0 \\
 \mulowergenerationnom \left( \capacityGeneration - \capacityGenerationLower \right) =0 \\
\end{align}

The CAPEX paybacks for generators %and transmission lines
\cref{eq:capital_price_generation_sum,eq:capital_price_flow_sum} change to 
\begin{align}
\pdv{\lagrangian}{\capacityGeneration}  = 0 \,\, \rightarrow \,\, \\
\capexGeneration =  \sum_t \muuppergeneration \, \generationpotential + \mulowergenerationnom - \muuppergenerationnom \Forall{n,s}
\label{eq:capex_generation_duality_bf}
\end{align}
for generators. % and 
% \begin{align}
% \pdv{\lagrangian}{\capacityFlow}  = 0 \,\, \rightarrow \,\, \\
% \capexFlow =  \sum_t \left( \muupperflow - \mulowerflow \right)  + \mulowerflownom - \muupperflownom \Forall{n,s}
% \label{eq:capex_flow_duality_bf}
% \end{align}
% for transmission lines.
Multiplying \cref{eq:capex_generation_duality_bf} by $ \capacityGeneration$ 
\begin{align}
 \capexGeneration \, \capacityGeneration =  \sum_t \muuppergeneration \, \generation +  \mulowergenerationnom \capacityGenerationLower - \muuppergenerationnom \capacityGenerationUpper \Forall{n,s}
\end{align}
Multiplying \cref{eq:capex_generation_duality_bf} by the amount of expansion $\left( \capacityGeneration - \capacityGenerationLower \right) $
\begin{align}
  \capexGeneration \, \capacityGeneration^{exp} =  \sum_t \muuppergeneration \, \left( \generation - \generationpotential \, \capacityGenerationLower \right) + \muuppergenerationnom \left( \capacityGenerationUpper - \capacityGenerationLower \right) \Forall{n,s}
\end{align}

playaround

\begin{align}
\capexGeneration \, \capacityGeneration =  \sum_t \muuppergeneration \, \generation +  \mulowergenerationnom \capacityGeneration - \muuppergenerationnom \capacityGeneration \Forall{n,s}
\end{align}
